---
title: "Exam 1 visualization"
format: html
---

```{r, echo=FALSE, message=FALSE}
#load packages
library(tidyverse)
library(tidytuesdayR)
library(gplots)
```

```{r, echo=FALSE, message=FALSE}
#get data
tuesdata <- tt_load('2020-02-18')
fc <- tuesdata$food_consumption
```

```{r}
#| fig-width: 13
#understand variables individually
ggplot(fc, aes(x = country)) +
  geom_bar()

ggplot(fc, aes(x = consumption)) +
  geom_density()

ggplot(fc, aes(x = food_category)) +
  geom_bar()
```

### For the variables of Food Category and Country, we don't learn too much other than each variable has an equal number of observations. However, the density plot of consumption provides a range of the consumption leves as well as the overall trend which average consumpotion being aroun 30 kg per person a year.

```{r}
#consumption by food category
ggplot(fc, aes(x = consumption, fill = food_category)) +
  geom_histogram() +
  scale_fill_viridis_d()
```
### Wheat and beef prodtucts tend to have some of the highest consumption through out the whole world

```{r}
#| fig-height: 22
#| fig-width: 15

ggplot(fc, aes(x = consumption))+
  geom_histogram() +
  facet_wrap(~ country)
```

### Western and "first world" country tend to have higher consumption than other countries

```{r}
#| fig-height: 20
#| fig-width: 20
#| fig-cap: "Food Consumption by country by food category (Zach Marshak, March 2025, data source: Nu3"
#| fig-alt: "Histograms, with the x axis as consumption by kilograms per person per year, facted by country, and filled by food category(Beef, eggs, fish, lam & goat, Dairy, Prok, poultry, rice, soybeans, and wheat)"

ggplot(fc, aes(x = consumption, fill = food_category)) +
  geom_histogram(position = "fill") +
  facet_wrap(~ country) +
  scale_fill_viridis_d() +
  labs(x = "Consumption", title = "Conusumption Throughout the World", fill = "Food Category") +
  theme(element_text(size = 12))
```

-  With so much data, it is really hard to tell, but by food category, consumption really depends on how a typical diet in each country is, and I still think the previous observation that Western countries typically have higher overall consumption holds true

#### Additonal questions

- What does Co2 emissions look like by food category in each country?
- What do Co2 emissions by consumption of each food category look like?

