---
title: "Food Consumption 2"
format: html
---

```{r, echo=FALSE, message=FALSE}
#install packages
library(tidytuesdayR)
library(tidyverse)
library(rnaturalearth)
library(sf)
```

```{r, echo=FALSE, message=FALSE}
#load data
tuesdata <- tt_load('2020-02-18')
fc <- tuesdata$food_consumption
```

```{r}
#fix food_cat vairable
fcc <- fc |>
  mutate(food_category = fct_recode(food_category,
                                    "Lamb" = "Lamb & Goat",
                                    "Dairy" = "Milk - inc. cheese",
                                    "Wheat" = "Wheat and Wheat Products",
                                    "Nuts" = "Nuts inc. Peanut Butter"))
```

```{r}
fcc |>
  distinct(food_category)
```

#### Research Question 1: What five countries consume the most food?

```{r}
fcc |>
  group_by(country) |>
  summarise(tot_consumption = sum(consumption)) |>
  arrange(desc(tot_consumption)) |>
  head(5)
```

```{r}
#| fig-height: 15
#| fig-width: 10
fcc |>
  group_by(country) |>
  summarise(tot_consumption = sum(consumption)) |>
  arrange(desc(tot_consumption)) |>
  ggplot(aes(x = tot_consumption, y = country)) +
           geom_col()
```

##### Research Question 2: What are the top 5 consuming countries by food category?

```{r}
top_5_by_cat <- fcc |>
  group_by(food_category) |>
  slice_max(order_by = consumption, n = 5)

print(top_5_by_cat)
```

```{r}
#| fig-height: 5
#| fig-width: 10
ggplot(top_5_by_cat, aes(x = food_category, y = consumption, fill = country)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d()
```

#### Research Question 3: What does the consumption of each food look like?


I copied the code from the test and kept getting errors on my end, so to improve it, I would apply a color-blind friendly scale using the scale_fill_virdis_c function, I would also apply an overall title for the plot and apply an alternative text using fig-cap alt
