---
title: "Solo Project"
format: html
---

```{r}
#install packages
library(tigris)
library(sf)
library(tidyverse)
```

```{r}
# get mn counties shapefile
MN <- counties(state = "Minnesota")
```

```{r}
# get labor force projections
labor_proj_bycounty <- read_csv("~/Documents/Github/portfolio-zmarshak/data/mn-counties_labor_rates.csv")
```

```{r}
#clean labor
labor_clean <- labor_proj_bycounty |>
  select(Geography, Age, `2015`, `2020`, `2025`, `2030`) |>
  group_by(Geography) |>
  summarise(`2015` = sum(`2015`), `2020` = sum(`2020`), `2025` = sum(`2025`), `2030` =  sum(`2030`)) |>
  mutate(labor_change = `2030` - `2020`)
```

```{r}
#| fig-height: 10
#| fig-width: 15
#| fig-alt: "A choropleth map of the change in Labor Force participation from 2020 to 2030 in Minnesota by county. The Twin Cities metro area is expected to see large growth over this period, and the rest of the state is expected to see much smaller growth or even negative growth."

MN |>
  select(NAME, geometry) |>
  left_join(labor_clean, join_by(NAME == Geography)) |>
  ggplot()+
  geom_sf(aes(fill = labor_change)) +
  scale_fill_viridis_c(direction = -1) +
  labs(title = "Minnesota Projected Labor Force Participation from 2020 to 2030", fill = "Projected Change by County", caption = "Projected Labor Force Participation Changes in Minnesota from 2020 to 2030 by county. (Created by: Zach Marshak April 2024, Data Sources: US Census Bureau, MN State Demographic Center)") +
  theme_void()
```

## Explaination 
In this visualization, I pulled labor force data from the Minnesota State Demographic Center. When visualized, you can see that the majority of counties in the state are expected to see very minimal growth and even shrinking labor force participation however, the counties in the Twin Cities Metro area are expected to see a much larger growth. This could because of growing populations in Twin Cities while there may likely be shrinking populations in the rural areas of the state.
